---
title: "Homework 1 (part 1)"
author: "Michael Pena"
date: "2024-01-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tinytex)
library(rgl)
```

## Problem 1

#i.

```{r}
# build f(x1,x2) function
f <- function(x1,x2){
  cos(x1*x2)
}
```

#ii.

```{r}
# build h(x1,x2) function
h <- function(x1,x2){
  1 - pi^2*(x1^2)/8
}
```

#iii.

```{r}
# build the sequences
x <- seq(-pi/4,pi/4,length = 30)
y <- seq(pi/4,3*pi/4,length = 30)
#grid <- expand.grid(x = x, y = y)

#fxy <- f(grid$x,grid$y)
#hxy <- h(grid$x,grid$y)
```


```{r}
# plot the functions
fsurf <- outer(x,y,FUN = f)
hsurf <- outer(x,y,FUN = h)
persp3d(x, y, fsurf, col = "red",shade = 0.3, alpha = 0.5)
persp3d(x, y, hsurf, col = "lightblue", shade = 0.3, alpha = 0.5, add=T)
rglwidget(controllers = ) 
```

#iv.

now we need to add subscripts

```{r}
#adding labels
persp3d(x, y, fsurf, col = "red",shade = 0.3, alpha = 0.5,
        main = "Taylor approximation of cos(x1x2)",
        xlab = "x1",
        ylab = "x2",
        zlab = "f")
persp3d(x, y, hsurf, col = "lightblue", shade = 0.3, alpha = 0.5, add=T)
rglwidget(controllers = ) 
```

#v. 

```{r}
#build error function
e <- function(x1,x2) {
  abs(f(x1,x2) - h(x1,x2))
}
```

```{r}
# render the plot of the function
esurf <- outer(x,y,FUN = e)
persp3d(x, y, esurf, col = "red",shade = 0.3, alpha = 0.5)
rglwidget(controllers = ) 
```

# vi.

```{r}
# add labelings
persp3d(x, y, esurf, col = "red",shade = 0.3, alpha = 0.5,
        main = "The error in second order Taylor expansion of cos (x1x2)",
        xlab = "x1",
        ylab = "x2",
        zlab = "e")
rglwidget(controllers = ) 
```

# vii. 

```{r}
# rendering contour for error function
filled.contour(x, y, esurf,
               xlab = "x1",
               ylab = "x2")
```

$h(x_1,x_2)$ estimates $f(x_1,x_2)$ closely where $x_1=0$. as we fall farther away from this region, $h$ and $f$ they diverge from eachother. This is clear by how the contour are gathered into the corners of the graphic. 
There is also a low contour in the around the line $x_2 = 1.7$; this is because the two surfaces cross each other at this curve. It would perhaps be best to use this Taylor approximation only around points that are around the line $x_1$.

## Problem 2

# (a).

$$
[\textbf{x} - \boldsymbol{\mu}]^T\boldsymbol{\Sigma}^{-1}[\textbf{x} - \boldsymbol{\mu}] 
= \sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{21})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2
$$

$$
f(\textbf{x}_0) = f\begin{pmatrix}
    \mu_1 \\
    \mu_2 
\end{pmatrix}
 = \frac{1}{2\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
$$


$$
(\textbf{x} - \textbf{x}_0)^T 
= \begin{pmatrix}
    x_1 - \mu_1  \\
    x_2 - \mu_2 
\end{pmatrix}
$$

$$
f_{x_1} = \frac{-e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{21})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)/(\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21})}}{4\pi (\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21})^{3/2}}
\cdot (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))
$$


$$
f_{x_2} = \frac{-e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{21})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)/(\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21})}}{4\pi (\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21})^{3/2}}
\cdot (2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{22})(x_1-\mu_1))
$$

$$
f_{x_1x_1} =\frac{e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{22})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\cdot \biggl( \frac{1}{2} (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))^2 - 2\sigma_{22}
\biggl)
$$

$$
f_{x_1x_2} = f_{x_2x_1} = \frac{e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{22})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\cdot \biggl( \frac{1}{2} (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))((2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{22})(x_1-\mu_1)) + \sigma_{12} + \sigma_{21}
\biggl)
$$

$$
f_{x_2x_2} = \frac{e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{22})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\cdot \biggl( \frac{1}{2} (2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{21})(x_1-\mu_1))^2 - 2\sigma_{11}
\biggl)
$$

$$
\nabla f\begin{pmatrix}
    \mu_1 \\
    \mu_2 
\end{pmatrix}
=
\frac{-e^0}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\begin{bmatrix}
    2\sigma_{22}(0)-(\sigma_{12}+\sigma_{22})(0) \\
    2\sigma_{11}(0)-(\sigma_{12}+\sigma_{22})(0)
\end{bmatrix}=
\frac{-1}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\begin{bmatrix}
    0 \\
   0
\end{bmatrix}=
\begin{bmatrix}
    0 \\
   0
\end{bmatrix}
$$
if 
$$
H\begin{pmatrix}
    \mu_1 \\
    \mu_2 
\end{pmatrix} =
\frac{-1}{4\pi (\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21})^{3/2}}
\begin{bmatrix}
    2\sigma_{22}  &
    -(\sigma_{12}+\sigma_{21})  \\
    -(\sigma_{12}+\sigma_{21}) &
    2\sigma_{11}
\end{bmatrix}
$$

our second order Taylor approximation about point $\begin{bmatrix}\mu_1\\ \mu_2 \end{bmatrix}$
$$
f(\textbf{x}) \cong 
\frac{1}{2\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}} - 
\frac{1}{8\pi (\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21})^{3/2}}
\biggl(2\sigma_{22}(x_1-\mu_1)^2 - 2(\sigma_{12} + \sigma_{21})(x_1-\mu_1)(x_2-\mu_2) + 2\sigma_{11}(x_2-\mu_2)^2 
\biggl) + R
$$


# (b).

```{r}
# building the function
f <- function(x1,x2,mu1,mu2,sig11,sig12,sig21,sig22){
  ((2*pi*sqrt(sig11*sig22 - sig21*sig12))^(-1))*exp((-0.5)*(2*sig22*(x1-mu1)^2 - (sig11*sig22 + sig12*sig21)*(x1-mu1)*(x2-mu2)+2*sig11*(x2-mu2)^2))
}

```

```{r}
#building the Taylor Series function
h <- function(x1,x2,mu1,mu2,sig11,sig12,sig21,sig22){
  ((2*pi*sqrt(sig11*sig22 - sig21*sig12))^(-1)) - ((8*pi*(sig11*sig22 - sig21*sig12)^(3/2))^(-1))*(2*sig22*(x1-mu1)^2 - 2*(sig12+sig21)*(x1-mu1)*(x2-mu2)+2*sig11*(x2-mu2)^2)
}
```

```{r}
# creating sequences
x <- seq(-1.5,1.5,length = 30)
y <- seq(-1.5,1.5,length = 30)
```

#### (i.)

```{r}
# render the surfaces
f_surf <- outer(x, y, mu1 = 0, mu2 = 0 ,sig11 = 1, sig12 = -0.3, sig21 = -0.3, sig22 = 1,FUN = f)
h_surf <- outer(x, y, mu1 = 0, mu2 = 0 ,sig11 = 1, sig12 = -0.3, sig21 = -0.3, sig22 = 1,FUN = h)
persp3d(x, y, f_surf, col = "red",shade = 0.7, alpha = 0.5)
persp3d(x, y, h_surf, col = "lightblue",shade = 0.7, alpha = 0.5, add = T)

rglwidget(controllers = ) 
```

#### (ii.)

```{r}
# render the surfaces
f_surf <- outer(x, y, mu1 = 0, mu2 = 0 ,sig11 = 1, sig12 = .8, sig21 = .8, sig22 = 1,FUN = f)
h_surf <- outer(x, y, mu1 = 0, mu2 = 0 ,sig11 = 1, sig12 = .8, sig21 = .8, sig22 = 1,FUN = h)
persp3d(x, y, f_surf, col = "red",shade = 0.7, alpha = 0.5)
persp3d(x, y, h_surf, col = "lightblue", shade = 0.7, alpha = 0.5, add=T)
rglwidget(controllers = ) 
```

### (c).

```{r}
# case (i.)
# rendering contour for error function
f_surf <- outer(x, y, mu1 = 0, mu2 = 0 ,sig11 = 1, sig12 = -0.3, sig21 = -0.3, sig22 = 1,FUN = f)
filled.contour(x, y, f_surf)
```

```{r}
# case (ii.)
# rendering contour for error function
f_surf <- outer(x, y, mu1 = 0, mu2 = 0 ,sig11 = 1, sig12 = .8, sig21 = .8, sig22 = 1,FUN = f)
filled.contour(x, y, f_surf)

```

The shape of this contour is an ellipses; the center is about the origin of the $x_1$-$x_2$ plane. As we approach the center, we climb higher in our z-axis.

### (d).

```{r}
# case i. eigenvectors
S1 <- matrix(c(1,-0.3,-0.3,1), nrow=2, ncol = 2)
# compute e-vectors and e-values
eigen(S1)
# superimpose the eigenvectors and eigenvalues over the contour
contour(x, y, f_surf)
arrows(0, 0, eigen(S1)$vectors[1, 1], eigen(S1)$vectors[2, 2], col = "green")
arrows(0, 0, eigen(S1)$vectors[2, 1], eigen(S1)$vectors[1, 2], col = "green")
```

```{r}
# case ii. eigenvectors
S2 <- matrix(c(1,0.8,0.8,1), nrow=2, ncol = 2)
# compute e-vectors and e-values
eigen(S2)
# superimpose the eigenvectors and eigenvalues over the contour
contour(x, y, f_surf)
arrows(0, 0, eigen(S2)$vectors[1, 1], eigen(S2)$vectors[2, 2], col = "green")
arrows(0, 0, eigen(S2)$vectors[2, 1], eigen(S2)$vectors[1, 2], col = "green")
```

eigenvectors and eigenvalues will typically correspond with the shape of the ellipses that are formed in the contour graphs.
