---
title: "Homework 1 (part 1)"
author: "Michael Pena"
date: "2024-01-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tinytex)
library(rgl)
```

## Problem 1

#i.

```{r}
# build f(x1,x2) function
f <- function(x1,x2){
  cos(x1*x2)
}
```

#ii.

```{r}
# build h(x1,x2) function
h <- function(x1,x2){
  1 - pi^2*(x1^2)/8
}
```

#iii.

```{r}
# build the sequences
x <- seq(-pi/4,pi/4,length = 30)
y <- seq(pi/4,3*pi/4,length = 30)
#grid <- expand.grid(x = x, y = y)

#fxy <- f(grid$x,grid$y)
#hxy <- h(grid$x,grid$y)
```


```{r}
# plot the functions
fsurf <- outer(x,y,FUN = f)
hsurf <- outer(x,y,FUN = h)
persp3d(x, y, fsurf, col = "red",shade = 0.3, alpha = 0.5)
persp3d(x, y, hsurf, col = "lightblue", shade = 0.3, alpha = 0.5, add=T)
rglwidget(controllers = ) 
```

#iv.

now we need to add subscripts

```{r}
#adding labels
persp3d(x, y, fsurf, col = "red",shade = 0.3, alpha = 0.5,
        main = "Taylor approximation of cos(x1x2)",
        xlab = "x1",
        ylab = "x2",
        zlab = "f")
persp3d(x, y, hsurf, col = "lightblue", shade = 0.3, alpha = 0.5, add=T)
rglwidget(controllers = ) 
```

#v. 

```{r}
#build error function
e <- function(x1,x2) {
  abs(f(x1,x2) - h(x1,x2))
}
```

```{r}
# render the plot of the function
esurf <- outer(x,y,FUN = e)
persp3d(x, y, esurf, col = "red",shade = 0.3, alpha = 0.5)
rglwidget(controllers = ) 
```

# vi.

```{r}
# add labelings
persp3d(x, y, esurf, col = "red",shade = 0.3, alpha = 0.5,
        main = "The error in second order Taylor expansion of cos (x1x2)",
        xlab = "x1",
        ylab = "x2",
        zlab = "e")
rglwidget(controllers = ) 
```

# vii. 

```{r}
# rendering contour for error function
filled.contour(x, y, esurf,
               xlab = "x1",
               ylab = "x2")
```

$h(x_1,x_2)$ estimates $f(x_1,x_2)$ closely where $x_1=0$. as we fall farther away from this region, $h$ and $f$ they diverge from eachother. This is clear by how the contour are gathered into the corners of the graphic. 
There is also a low contour in the around the line $x_2 = 1.7$; this is because the two surfaces cross each other at this curve. It would perhaps be best to use this Taylor approximation only around points that are around the line $x_1$.

## Problem 2

# (a).

\[
[\textbf{x} - \boldsymbol{\mu}]^T\boldsymbol{\Sigma}^{-1}[\textbf{x} - \boldsymbol{\mu}] 
= \sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{21})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2
\\
f(\textbf{x}_0) = f\begin{pmatrix}
    \mu_1 \\
    \mu_2 
\end{pmatrix}
 = \frac{1}{2\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\\ 
(\textbf{x} - \textbf{x}_0)^T 
= \begin{pmatrix}
    x_1 - \mu_1  \\
    x_2 - \mu_2 
\end{pmatrix}
\\
f_{x_1} = \frac{-e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{21})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\cdot (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))
\\
f_{x_2} = \frac{-e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{21})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\cdot (2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{22})(x_1-\mu_1))
\\
f_{x_1x_1} =\frac{e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{22})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\cdot \biggl( \frac{1}{2} (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))^2 - 2\sigma_{22}
\biggl)
\\
f_{x_1x_2} = f_{x_2x_1} = \frac{e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{22})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\cdot \biggl( \frac{1}{2} (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))((2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{22})(x_1-\mu_1)) + \sigma_{12} + \sigma_{21}
\biggl)
\\
f_{x_2x_2} = \frac{e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{22})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\cdot \biggl( \frac{1}{2} (2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{21})(x_1-\mu_1))^2 - 2\sigma_{11}
\biggl)
\\
\nabla f\begin{pmatrix}
    \mu_1 \\
    \mu_2 
\end{pmatrix}
=
\frac{-e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{22})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\begin{bmatrix}
    2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{22})(x_2-\mu_2) \\
    2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{22})(x_1-\mu_1)
\end{bmatrix}
\\
H(\textbf{x}_0) =
\frac{e^{\frac{-1}{2}(\sigma_{22}(x_1-\mu_1)^2 - (\sigma_{12}+\sigma_{22})(x_1-\mu_1)(x_2-\mu_2) + \sigma_{11}(x_2-\mu_2)^2)}}{4\pi \sqrt{\sigma_{11}\sigma_{22}-\sigma_{12}\sigma_{21}}}
\begin{bmatrix}
    \frac{1}{2} (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))^2 - 2\sigma_{22} &
    \frac{1}{2} (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))((2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{22})(x_1-\mu_1)) + \sigma_{12} + \sigma_{21} \\
    \frac{1}{2} (2\sigma_{22}(x_1-\mu_1)-(\sigma_{12}+\sigma_{21})(x_2-\mu_2))((2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{22})(x_1-\mu_1)) + \sigma_{12} + \sigma_{21} &
     \frac{1}{2} (2\sigma_{11}(x_2-\mu_2)-(\sigma_{12}+\sigma_{21})(x_1-\mu_1))^2 - 2\sigma_{11}
\end{bmatrix}
\\
\]

# (b).

\[

\]$

