---
title: 'Exam #1'
author: "Michael Pena"
date: "2024-06-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data <- read.table("TotalSales-2.txt",header = T)
series <- ts(data,frequency = 4)
library(dplyr)
```

# Concepts and Theoretical

### 1.

```{r}
set.seed(536.1)
# initialize X
sample(c(-1,1),size = 500,prob = c(.5,.5),replace = T) -> X
# initialize Svec 
Svec = cumsum(c(0,X))
# make graphics
plot.ts(Svec,type = "l",ylab = "S_t", xlab = "t")
acf(Svec)
```


```{r}
# running means
Svec_bar = Svec
for(i in 1:501){
  mean(Svec[1:i]) -> Svec_bar[i]
}
# plot running means
plot(Svec_bar,type = 'l')
```
 Above graph showing that the expected value will change depending on time. System is not stationary.

### 2. 

find $\mu$
because system is stationary, we know that $E[X_t] = E[X_{t-1}] = E[X_{t-2}] = \mu$
but since 
$$
E[X_t] = E[0.2X_{t-1} + 0.4X_{t-2} + Z_t]
\\E[X_t] = E[0.2X_{t-1} + 0.4X_{t-2} ] + 0
\\E[X_t] =0.2E[X_{t-1}] + 0.4E[X_{t-2}]
\\\mu = 0.6\mu
$$
this can only be true if $\mu=0$

$$
X_t = 0.2X_{t-1} + 0.4X_{t-2} + Z_t\\
\Rightarrow
X_{t-h}X_t = 0.2X_{t-1}X_{t-h} + 0.4X_{t-2}X_{t-h} + Z_tX_{t-h}\\
\Rightarrow
E[X_{t-h}X_t] = E[0.2X_{t-1}X_{t-h} + 0.4X_{t-2}X_{t-h} + Z_tX_{t-h}]\\
= 0.2E[X_{t-1}X_{t-h}] + 0.4E[X_{t-2}X_{t-h}] + E[Z_tX_{t-h}]\\
\Rightarrow
E[X_{t-h}X_t]= 0.2E[X_{t-1}X_{t-h}] + 0.4E[X_{t-2}X_{t-h}]\\
\Rightarrow \gamma(h) = 0.2\gamma(h-1)+0.4\gamma(h-2)\\
\Rightarrow 0 = 0.2\gamma(h-1)+0.4\gamma(h-2) - \gamma(h)
$$

knowing $\rho_X(h) = \frac{\gamma(h)}{\gamma(0)}$ divide the last line all by $\gamma(0)$

$$
0 = 0.2\rho_X(h-1)+0.4\rho_X(h-2) - \rho_X(h)
$$

knowing this system is stationary than we know that $\rho_X(0)=1$, set $h=1$.

$$
0 = 0.2\rho_X(0)+0.4\rho_X(-1) - \rho_X(1)\\
\to 0 = 0.2+0.4\rho_X(-1) - \rho_X(1)\\
\to 0 = 0.2 - 0.6\rho_X(\pm1)\\
\to \frac{1}{3} = \rho_X(\pm1)\\
\rho_X(h) = \begin{cases}
  1 & h = 0 \\
  1/3 & |h| = 1 \\
  0.2\rho_X(h-1) + 0.4\rho_X(h-2) & |h| \ge 2
\end{cases}
$$



### 3(a).






# Data Analysis

```{r}
plot.ts(series,type = 'l',ylab = "series")
abline(lm(series ~ time(series),data = data))
plot.ts(log(series),type = "l",ylab="log of series")
abline(lm(log(series) ~ time(series),data = data))
```

```{r}
#define a t and cycle
t = time(series)
cyc = as.factor(cycle(series))
```


```{r}
# test two regressions 
fit1 <- lm(log(series) ~ cyc + t)
fit2 = lm(log(series) ~ (cyc*t)^3 + cyc*t+ cyc)
plot.ts(log(series),type = "l",ylab="log of series")
points(t,predict.lm(fit1),type = "l", col = "red" )
points(t,predict.lm(fit2),type = "l", col = "blue" )

```


```{r}
# diagnostics for fit 1
par(mfrow=c(2,2)) # Dividing the plotting page into 4 panels
plot(fit1$fitted, fit1$residuals) # plot of fitted values vs residuals 
qqnorm(fit1$residuals) #qq-plot of residuals
qqline(fit1$residuals) # plotting the line, along which the dots in qq-plot should lie 
plot(fit1$residuals) # plotting the residuals vs time
abline(h=0,lty=2) # plotting a horizontal line at 0
acf(fit1$residuals) #sample acf plot of residuals

# diagnostics for fit 1
par(mfrow=c(2,2)) # Dividing the plotting page into 4 panels
plot(fit2$fitted, fit2$residuals) # plot of fitted values vs residuals 
qqnorm(fit2$residuals) #qq-plot of residuals
qqline(fit2$residuals) # plotting the line, along which the dots in qq-plot should lie 
plot(fit2$residuals) # plotting the residuals vs time
abline(h=0,lty=2) # plotting a horizontal line at 0
acf(fit2$residuals) #sample acf plot of residuals
```



Our second model seems to fair better in the diagnostics

```{r}
# get MSE for both
MSE <- function(fit,yhat){
  y <- predict.lm(fit)
  dif <- abs(y - yhat)
  mse <- (1/length(y)) * sum(dif^2)
  return(mse)
}

# get AIC/BIC for both models
ABIC <- function(fit){
  n <- length(predict.lm(fit))
  k <- length(fit$coefficients) - 1
  L <- as.numeric(logLik(fit))
  a <- 2*k - log(L)*2
  b <- k*log(n) - 2*log(L)
  return(list(AIC = a, BIC = b))
}
```

```{r}
sprintf("first model's MSE is %f while the second model has an MSE of %f",MSE(fit1,series),MSE(fit2,series))
sprintf("first model AIC is %f while BIC is %f", ABIC(fit1)$AIC, ABIC(fit1)$BIC)
sprintf("second model AIC is %f while BIC is %f", ABIC(fit2)$AIC, ABIC(fit2)$BIC)
```
T
