---
title: "kidneystone"
output: pdf_document
date: "2024-01-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE}
# loading packages
library(caret)
library(yardstick)
library(dplyr)
library(modelr)
library(ggplot2)
library(GGally)

```

```{r}
# loading data into dataframes
stone <- read.csv("kindey stone urine analysis.csv")
```

```{r}
#split data for train and testing 
set.seed(530)

#building a new column called "train"
stone <- stone %>% mutate(train = FALSE) 
trainIndex <- createDataPartition(stone$target, p = 0.7, list = FALSE, times = 1)
stone$train[trainIndex] <- TRUE

#rename the new test a train sets
stone_train <- stone %>%  filter(train == TRUE)
stone_test <- stone %>% filter(train == FALSE)
```


```{r}
stone_train$target <- as.character(stone_train$target)
```



```{r message=FALSE}
ggpairs(stone_train,aes(color = target))
```
gravity
specific gravity of urine

ph
ph of urine

osmo
osmolarity of urine

cond
conductivity of urine

urea
concentration of urea in urine

calc
concentration of calcium in urine

target
0- absence of stone 1- presence of stone


## our new math model
Y : target
X_i's : gravity + osmo + cond + urea + calc

## our math model to copy+paste
target ~ gravity + osmo + cond + urea + calc

# train data with stone_train set 

```{r}
# train a logistic model
fit_logit <- train(target ~ gravity + osmo + cond + urea + calc,
                   data = stone_train,
                   method = "glm",
                   family = "binomial",
                   trControl = trainControl(method = 'LOOCV'))
```

```{r}
# train a naives-bayes model
fit_nb <- train(target ~ gravity + osmo + cond + urea + calc,
                data = stone_train,
                method = "naive_bayes")
```

```{r}
# train a LDA model
fit_lda <- train(target ~ gravity + osmo + cond + urea + calc,
                 data = stone_train,
                 method = "lda")
```

```{r}
# train a QDA model
fit_qda <- train(target ~ gravity + osmo + cond + urea + calc,
                 data = stone_train,
                 method = "qda")
```


# test with the test set

```{r}
# making new columns on the test set
stone_test  <- stone_test %>% 
  mutate(logit_yhat = predict(fit_logit,newdata=.,type="raw")) %>%
  mutate(logit_prob = predict(fit_logit, newdata = .,type="prob")$'1')
```

```{r}
# making even more new columns on the test set

# columns for naive_bayes predictions
stone_test <- stone_test %>%
  mutate(nb_yhat = predict(fit_nb, newdata =., type = "raw")) %>%
  mutate(nb_prob = predict(fit_nb, newdata=., type = "prob")$'1')

# columns for LDA predictions
stone_test <- stone_test %>%
  mutate(lda_yhat = predict(fit_lda, newdata=., type="raw")) %>%
  mutate(lda_prob = predict(fit_lda, newdata=., type="prob")$'1')

# columns for QDA predictions
stone_test <- stone_test %>%
  mutate(qda_yhat = predict(fit_qda, newdata=., type = "raw")) %>%
  mutate(qda_prob = predict(fit_qda, newdata=., type = "prob")$'1')
```


```{r}
stone_test$target <- as.factor(stone_test$target)

stone_test %>% accuracy(truth=target,estimate=logit_yhat)
  stone_test %>% accuracy(truth = target, estimate = nb_yhat)
  stone_test %>% accuracy(truth = target, estimate = lda_yhat)
  stone_test %>% accuracy(truth = target, estimate = qda_yhat)
```

# make confidence matrices
```{r}
# for logistic
stone_test %>% conf_mat(target,logit_yhat)
# for naive-bayes
stone_test %>% conf_mat(target,nb_yhat)
#for lda
stone_test %>% conf_mat(target,lda_yhat)
# for qda
stone_test %>% conf_mat(target,qda_yhat)
```

# make ROC curve
```{r}
#ROC for logistic 
stone_test %>% 
  roc_curve(truth = target,
            estimate = logit_prob,
            event_level = "second") %>%
  autoplot()


# ROC for naive-bayes
stone_test %>% 
  roc_curve(truth = target,
            estimate = nb_prob,
            event_level = "second") %>%
  autoplot()

# ROC for LDA
stone_test %>% 
  roc_curve(truth = target,
            estimate = lda_prob,
            event_level = "second") %>%
  autoplot()

# ROC for QDA
stone_test %>% 
  roc_curve(truth = target,
            estimate = qda_prob,
            event_level = "second") %>%
  autoplot()
```

# compute area under curve
```{r}
#for logistic 
stone_test %>% roc_auc(target,logit_prob, event_level = 'second')
# for nai-bayes
stone_test %>% roc_auc(target,nb_prob, event_level = 'second')
# for lda
stone_test %>% roc_auc(target,lda_prob, event_level = 'second')
# for qda 
stone_test %>% roc_auc(target,qda_prob, event_level = 'second')
```


I want to go with QDA because it has the highest accuracy but the second highest AUC of the ROC. 

I did not pick logistic because while it had the largest AUC of the ROC, it had the lowest accuracy.