---
title: "Homework 1"
subtitle: "computational"
author: "Michael Pena"
format: pdf
editor: visual
---

#### Question 2

(a).

|   $$               
            X_1      
                $$   |                   $$                                                                       
                                                                                                 P(X_1)           
                                                                                                            $$    |
|:----------------:|:----------------------------------------------------:|
|         0          |                                      .10 + .15 = .25                                       |
|         1          |                                        .2 + .1 = .3                                        |
|         2          |                                       .15 + .3 = .45                                       |

(b).

|   $$               
            X_2      
                $$   |                   $$                                                                           
                                                                                                   P(X_2)             
                                                                                                                $$    |
|:----------------:|:----------------------------------------------------:|
|         0          |                                      .1 + .2 + .15 = .45                                       |
|         1          |                                      .15 + .1 + .3 = .55                                       |

(c).

|   $$               
            X_1      
                $$   |                  $$                                                            
                                                                                             X_2      
                                                                                                 $$   |                                                                $$                                                                                                                                                                           
                                                                                                                                                                                                                                                                                                                                    P(X_1|X_2)      
                                                                                                                                                                                                                                                                                                                                              $$    |
|:----------------:|:-----------------:|:--------------------------------:|
|         0          |                                       0                                        |                                                                                                                     2/9                                                                                                                     |
|         1          |                                       0                                        |                                                                                                                     4/9                                                                                                                     |
|         2          |                                       0                                        |                                                                                                                     1/3                                                                                                                     |
|         0          |                                       1                                        |                                                                                                                   0.27272                                                                                                                   |
|         1          |                                       1                                        |                                                                                                                   0.1818                                                                                                                    |
|         2          |                                       1                                        |                                                                                                                   0.5454                                                                                                                    |

(d).

|   $$               
            X_1      
                $$   |                  $$                                                            
                                                                                             X_2      
                                                                                                 $$   |                                                                $$                                                                                                                                                                           
                                                                                                                                                                                                                                                                                                                                    P(X_2|X_1)      
                                                                                                                                                                                                                                                                                                                                              $$    |
|:----------------:|:-----------------:|:--------------------------------:|
|         0          |                                       0                                        |                                                                                                                     0.4                                                                                                                     |
|         1          |                                       0                                        |                                                                                                                     2/3                                                                                                                     |
|         2          |                                       0                                        |                                                                                                                     1/3                                                                                                                     |
|         0          |                                       1                                        |                                                                                                                     0.6                                                                                                                     |
|         1          |                                       1                                        |                                                                                                                     1/3                                                                                                                     |
|         2          |                                       1                                        |                                                                                                                     2/3                                                                                                                     |

\(e\)

It's not independent because $P(X_2 = 0|X_1 = 0) = 0.4$ and $P(X_2=0) = 0.45$ which are not the same.

#### Question 3

```{r}
yi = c(1.8,1.7,1.4,1.6,1.9,1.5,1.2)
```

(a).

$$
\frac{\partial}{\partial \beta} \sum_{i=1}^{7} ln(\frac{\beta^4}{6}y_i^3e^{-\beta y_i}) 
\\=
 \sum_{i=1}^{7} \Big( \frac{\beta^4}{6}y_i^3e^{-\beta y_i} \Big)^{-1}
\cdot y_i^3 \Big( \frac{2\beta^3}{3}e^{-\beta y_i}-\frac{\beta^4}{6}y_ie^{-\beta y_i} \Big)
\\=
\sum_{i=1}^{7} \frac{6}{\beta}(\frac{2}{3} - \frac{\beta y_i}{6})
\\= 
\frac{4(7)}{\beta} - \sum_{i=1}^7 y_i
$$

this implies that $\tilde{\beta} = \frac{28}{\sum_{i=1}^7 y_i}=2.522523$

(b).

solving for $\tilde\alpha$

$$
\frac{\partial}{\partial \alpha} \Bigg[ \sum_{i=1}^7 \Big(
\alpha ln2 - ln(\Gamma(\alpha)) + (\alpha-1)ln(y_i)-2y_i
\Big)\Bigg]
\\=
\sum_{i=1}^7 \Big(
ln2 - \frac{\Gamma(\alpha)\psi(\alpha)}{\Gamma(\alpha)} + ln(y_i)
\Big)
\\=
7ln2 - \psi(\alpha) + \sum^{7}_{i=1}ln(y_i)
$$

so if we set the derivative to zero, it implies

$$
\psi(\tilde\alpha) = 7ln2 + \sum^{7}_{i=1}ln(y_i)
$$

furthermore

$$
\tilde\alpha = \psi^{-1}\Big( 7ln2 + \sum^{7}_{i=1}ln(y_i)\Big)
$$

I am assuming this is this non closed ended issue that was mentioned in the handout. Let's graph the log-likelihood instead.

```{r}

LLH <- function(a, y = yi){
  n = length(y)
  f = n*log(2^a/gamma(a)) + (a - 1)*log(prod(y)) - 2*sum(y)
  return(f)
}

alpha <- seq(0,10,length = 1000)
plot(alpha,LLH(alpha),type = 'l')
```

```{r}
alpha <- seq(3.60,3.70,length = 1000)
plot(alpha,LLH(alpha),type = "l")
abline(v=3.625,col = "red")
```

Eyeballing this to be about $\tilde\alpha = 3.625$

#### Question 4

(a).

$$
P(\beta|y_i) \propto \Big(\prod_{i=1}^7\frac{\beta^4}{6}y_i^3e^{-\beta y_i} \Big) \Big(0.8e^{-0.8\beta} \Big)
$$

$$
= \frac{0.8}{6^7}(\prod_i^7 y_i^3)\beta^{28} e^{-\beta(0.8 + \sum_i^7 y_)}
\\=
0.03681441\beta^{28} e^{-11.9\beta}
$$

```{r}
0.8/6^7*prod(yi^3) -> c1
0.8+sum(yi) -> c2
c1;c2
```

$$
P(\beta|y_i) = \frac{0.03681441\beta^{28} e^{-11.9\beta}}{\int_{0}^{\infty} 0.03681441\beta^{28} e^{-11.9\beta} d\beta}
$$

(b).

```{r}
# plotting the posterior
post <- function(beta){
 # numerator
  numer <- function(x){
    0.03681441 * x^28 * exp(-c2*x)
  }
 # denominator
  denom <- integrate(numer, lower = 0, upper = Inf)$value
 # output
  return(numer(beta)/denom)
}
# render
beta <- seq(0,10,length = 1000)
plot(beta,post(beta),type = "l")
```

(c).

let's try an easy numerical evaluation approach

```{r}
# make vector between 2 and 3
beta1 <- seq(2,3,length = 500000)
beta1[which(post(beta1) == max(post(beta1)))] -> MLE_beta
MLE_beta
```

(d).

let's generate some beta values using sample()

```{r}
  # render some values of beta from 0 to 6
  beta_sampl = runif(1000000,0,6)
  # get p_values
  p_sampl = post(beta_sampl)
  B = sample(size = 5000, beta_sampl,replace = T, prob = p_sampl)
```

```{r}
hist(B,breaks = 80, freq = F, xlab = "beta")
lines(beta,post(beta),col = "forestgreen", lwd = 2)
```

```{r}
mean(B)
for(v in c(0.025,.5,.975)){
  print(quantile(B,v))
  }
```

```         
```
