---
title: "Titanic Guided Project"
subtitle: "Michael Peña"
format: html
editor: visual
---

## Data Processing

```{r}
library(dplyr)
library(caret)
library(forecast)
library(naivebayes)
library(MASS)
X <- read.csv("train.csv",header = T)
X <- X %>% dplyr::select(Survived,Pclass,Sex,Age,Fare,SibSp,Parch)
```

Imputing age

```{r}
meanAge = mean(X$Age[!is.na(X$Age)])
for(i in 1:length(X$Age)){
  if(is.na(X$Age[i]) == 1){X$Age[i] = meanAge}
}
```

We will make Sex numerical

```{r}
for(i in 1:891){
  if(X$Sex[i] == "male"){X$Sex[i] = 1}
  else{X$Sex[i] = 0}
}
X$Survived <- as.integer(X$Survived)
X$Sex <- as.integer(X$Sex)
```

splitting the data (notice 80% 0f 891 $\approx$ 713)

```{r}
set.seed(533)
trainIndex <- sample(1:891, size = 713, replace = F)
Xtrain <- X[trainIndex,]
Xtest <- X[-trainIndex,]
```

## Logistic Regression

fit to a logistic regression

```{r}
# make sure Survived is a factor
Xtrain$Survived <-as.factor(Xtrain$Survived) 
# train the logit
fitLogit <- train(Survived ~ .,
                  data= Xtrain,
                  method = "glm",
                  family = "binomial") 
# print summary
summary(fitLogit)
```

looking at the above summary, it seems as though Pclass and Sex have a major influence on survival.

```{r}
# doing this to get the accuracy
Xtest <- Xtest %>%
  mutate(logityhat = predict(fitLogit, newdata =., type = "raw"))
# confusion matrix
Xtest$Survived <- as.factor(Xtest$Survived)
confusionMatrix(Xtest$logityhat,Xtest$Survived)$table -> cm.log
# display for accuracy, precision, recall, f1score
table1 <- function(tab,name, printtnr = F){
  tab[1,1] -> tn
  tab[2,2] -> tp
  tab[1,2] -> fp
  tab[2,1] -> fn
  a = (tp+tn)/(tp+tn+fp+fn)
  p = tp/(tp+fp)
  r = tp/(tp+fn)
  f1 = 2*p*r/(p+r)
  tnr = tn/(tn+fp)
  if(printtnr == 0){  
    sprintf("%s || Accuracy: %f | Precision: %f | Recall(TPR): %f | F1: %f",name,a,p,r,f1)
  } else{
    sprintf("%s || Accuracy: %f | Precision: %f | Recall(TPR): %f | F1: %f | TNR: %f",name,a,p,r,f1,tnr)  
  }
  }

table1(cm.log,"Logistic Regression")
```

## Linear Discriminant Analysis (LDA) or Quadratic Discriminant Analysis (QDA)

let's fit a QDA

```{r}
# training on the QDA
fitQda <- (train(method = "qda", Survived ~ ., data = Xtrain))
# accuracy
Xtest <- Xtest %>%
  mutate(qdayhat = predict(fitQda, newdata =., type = "raw"))
# comparing accuracy
confusionMatrix(Xtest$logityhat,Xtest$Survived)$overall["Accuracy"]
confusionMatrix(Xtest$qdayhat,Xtest$Survived) -> cm.qda
cm.qda$overall["Accuracy"]
```

the QDA accuracy is better than that of the Logistic. Although this is strange as QDA assumes normality in the predictors and I didn't check for that.

```{r}
library(klaR)
partimat(formula = Survived ~ Age + Fare,data=Xtest,method="qda",main="QDA Partition Plots")

```

## Naive Bayes

```{r}
fitNb <- train(method = "naive_bayes", Survived ~ . , data = Xtrain)
Xtest <- Xtest %>%
  mutate(nbyhat = predict(fitNb, newdata =., type = "raw"))
confusionMatrix(Xtest$nbyhat,Xtest$Survived) -> cm.nb
cm.nb$overall["Accuracy"]

```

Naive Bayes has a similar accuracy to logistic regression. Let's see if certain variables improve it's performance metrics.

```{r}
pca <- prcomp(Xtrain[,2:7], scale. = T)
evar = pca$sdev^2 / sum(pca$sdev^2)
plot(evar, type = 'b')
```

looks like 3 components are relevant for our analysis, maybe we can see which ones

```{r}
pca$rotation[, 1:3]
```

rotation tells me that Fare is important in three components as it places in all three, so it must be important. SibSp places in two components, while Sex seems to be very influential in component three, so I will choose it.

```{r}
fitNb2 <- train(method = "naive_bayes", Survived ~ Fare + SibSp + Sex , data = Xtrain)
Xtest <- Xtest %>%
  mutate(nb2yhat = predict(fitNb2, newdata =., type = "raw"))
confusionMatrix(Xtest$nb2yhat,Xtest$Survived) -> cm.nb2
table1(cm.nb2$table,"Naive Bayes with less variables")
table1(cm.nb$table,"Naive Bayes")
```

slight improvement in accuracy and TNR, making it relatively better than all other models so far. Let's see if all other models improve after this variables reduction.

```{r}
# log regression
fitLogit <- train(Survived ~ Fare + SibSp + Sex ,
                  data= Xtrain,
                  method = "glm",
                  family = "binomial") 
Xtest$logityhat <- predict(fitLogit, newdata =Xtest, type = "raw")
confusionMatrix(Xtest$logityhat,Xtest$Survived)$table -> cm.log    

# QDA 
fitQda <- (train(method = "qda", Survived ~ Fare + SibSp + Sex , data = Xtrain))
Xtest$nbyhat <- predict(fitQda, newdata =Xtest, type = "raw")
confusionMatrix(Xtest$nbyhat,Xtest$Survived)$table -> cm.qda2

# performance metrics
table1(cm.log,"Logistic Regression with less variables",T)
table1(cm.qda2,"QDA with less variables",T)
table1(cm.nb2$table,"Naive Bayes with less variables",T)
```

Only using the three variables, Naive Bayes and Logistic Regression improved, however QDA seems to have taken a dive in performance considering the previous accuracy score was around .76.

## Model Comparison

```{r}
# performance metrics
table1(cm.log,"Logistic Regression with less variables",T)
table1(cm.qda$table,"QDA",T)
table1(cm.nb2$table,"Naive Bayes with less variables",T)
```

|                model                 | Accuracy | Precision | Recall(TPR) |    F1    |   TNR    |
|:-----------------:|:---------:|:---------:|:---------:|:---------:|:---------:|
| Logistic Regression (less variables) | 0.752809 | 0.649123  |  0.606557   | 0.627119 | 0.829060 |
|         QDA (all variables)          | 0.764045 | 0.736842  |  0.608696   | 0.666667 | 0.862385 |
|     Naive Bayes (less variables)     | 0.758427 | 0.666667  |  0.612903   | 0.638655 | 0.836207 |

QDA dominates in all of these performance metrics, including accuracy. It also has the highest true positive and and true negative rate; it is the best model for predicting who survived as well as who did not.

## Conclusion

Quadratic Discriminant Analysis with all the 6 variables is the model that has the most reliable results. If we are predicting who survived the Titanic, given these 6 particular variables from the data, then this the model I would move forward with.
