---
title: "e4MP_Submission"
author: "Michael Pena"
date: "2024-06-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Desktop/githubbahubba/CSUF/summer 2024/Exam4")
library(tseries)
library(forecast)
```

# Problem 1

### (a).

```{r}
# loading the data
data <- read.table("UKg2024.txt", head=T)
series.train <- ts(data[1:48,1],start = c(1980,1), frequency = 12)
series.val <- ts(data[49:60,1],start = c(1980,1), frequency = 12)
# plotting ts
plot.ts(series.train)
plot(decompose(series.train))
plot.ts(decompose(series.train)$season[1:12])
kpss.test(decompose(series.train)$random)
```

Our general trend snakes upwards with a some dips in 1981 and 1982. 
Seasonality is clear and repeats every year. There is a trough in the beginning of the year and two peaks on May and August.

Our KPSS test concludes that this system is most likely stationary.

### (b).

```{r}
#Note that our first month is actually July
X <- as.factor(c(rep(seq(1:12),4)))
cen.mdat <- c(scale(series.train, center = TRUE, scale = FALSE))
mod.lm <- lm(cen.mdat ~ X - 1)
# print out the summary
summary(mod.lm)
par(mfrow = c(2,2))
plot(mod.lm)
```


scale location isn't linear; we may have data that does not have constant variance. If this is the case we may need to difference our data in the future.

```{r}
par(mfrow = c(1,2))
acf(cen.mdat)
pacf(cen.mdat)
auto.arima(cen.mdat)
```

