---
title: "Homework 2 (Part 2)"
author: "Michael Pena"
date: "2024-02-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```

## (a).

General Secant Method

$$
x_{n+2} = x_{n+1} - f(x_{n+1})\frac{x_{n+1}-x_{n}}{f(x_{n+1})-f(x_{n})}
$$

$$
\ell'(\theta) = \frac{1997}{2+\theta} - \frac{1811}{1-\theta} + \frac{32}{\theta} 
$$

```{r}
# render derivative
f <- function(theta){
  1997/(2 + theta) - 1811/(1-theta) + 32/theta
}

# build secant method


secM <- function(theta0,theta1,maxit,tolerr,tolgrad){
# input initials
  it = 0
  absg <- abs(f(theta1))
  mre <- abs(theta1 - theta0)/max(1,abs(theta1))
  store <- data.frame()
  while(it < maxit && (absg > tolgrad || mre > tolerr)){
    # secant method equation
    theta2 <- theta1 - f(theta1)*(theta1 - theta0)/(f(theta1) - f(theta0))
    # storing into dataframe
    absg <- abs(f(theta2))
    mre <- abs(theta1 - theta0)/max(1,abs(theta1))

    # end of the loop
     theta0 <- theta1
     theta1 <- theta2
     it = it + 1
     row <- c(it,theta2,mre,absg)
     store <- rbind(store,row)
  }
# formatting dataframe
  store <- data.frame(store) %>% set_names("Iteration","Theta","Relative Error","Gradient at Theta") %>% mutate(
  Theta = sprintf("%12.12f",Theta),
  `Relative Error` = sprintf("%.1e", `Relative Error`),
  `Gradient at Theta` = sprintf("%.1e", `Gradient at Theta`)
  )  
 return(store)
}

```


```{r}
# start process
secM(theta0 = .02,.01,20,1e-6,1e-9)

```

## (b).

```{r}
secM2 <- function(theta0,theta1,maxit,tolerr,tolgrad,star){
# input initials
  it = 0
  gold = (1 + sqrt(5))/2
  absg <- abs(f(theta1))
  mre <- abs(theta1 - theta0)/max(1,abs(theta1))
  store <- data.frame()
  while(it < maxit && (absg > tolgrad || mre > tolerr)){
    # secant method equation
    theta2 <- theta1 - f(theta1)*(theta1 - theta0)/(f(theta1) - f(theta0))
    # storing into dataframe
    absg <- abs(f(theta2))
    mre <- abs(theta1 - theta0)/max(1,abs(theta1))
    rat <- abs(theta1-star)/(abs(theta1 - theta0))^(gold)
    sigdig <- -log10(abs(theta1 - star)/abs(star))

    # end of the loop
     theta0 <- theta1
     theta1 <- theta2
     it = it + 1
     row <- c(it,theta2,rat,sigdig)
     store <- rbind(store,row)
  }
# formatting dataframe
store <- data.frame(store) %>% set_names("Iteration","Theta","Convergence Ratio","Sig. Digits") %>% mutate(
  Iteration = sprintf("%02.f",Iteration),
  Theta = sprintf("%12.12f",Theta),
  `Convergence Ratio` = sprintf("%.3e",`Convergence Ratio`)
  )  
 return(store)
}
```


```{r}
thetstar <- (-1657 + sqrt(3728689))/7680
secM2(.02,.01,20,1e-6,1e-9,thetstar)
```

## (c).

```{r}
secM2.c <- function(theta0,theta1,maxit,tolerr,tolgrad,star){
# input initials
  it = 0
  gold = (1 + sqrt(5))/2
  absg <- abs(f(theta1))
  mre <- abs(theta1 - theta0)/max(1,abs(theta1))
  store <- data.frame()
  while(it < maxit && (absg > tolgrad || mre > tolerr)){
    # secant method equation
    theta2 <- theta1 - f(theta1)*(theta1 - theta0)/(f(theta1) - f(theta0))
    # storing into dataframe
    absg <- abs(f(theta2))
    mre <- abs(theta1 - theta0)/max(1,abs(theta1))
    ratsl <- abs(theta1-star)/(abs(theta1 - theta0))
    ratquad <- abs(theta1-star)/(abs(theta1 - theta0))^2
    sigdig <- -log10(abs(theta1 - star)/abs(star))

    # end of the loop
     theta0 <- theta1
     theta1 <- theta2
     it = it + 1
     row <- c(it,theta2,ratsl,ratquad,sigdig)
     store <- rbind(store,row)
  }
# formatting dataframe
store <- data.frame(store) %>% set_names("Iteration","Theta","Convergence Ratio Lin","Convergence Ratio Quad","Sig. Digits") %>% mutate(
  Iteration = sprintf("%02.f",Iteration), 
  Theta = sprintf("%12.12f",Theta)  
  )  
 return(store)
}
```

```{r}
thetstar <- (-1657 + sqrt(3728689))/7680
df.c <- secM2.c(.02,.01,20,1e-10,1e-12,thetstar)[,3:4]
df.c
plot(1:length(df.c[,1]),df.c[,1],type = 'b')
plot(1:length(df.c[,1]),df.c[,2],type = 'b')

```

this seems to converge linearly as 

>> ratsl <- abs(theta1-star)/(abs(theta1 - theta0))

approaches zero while 

>> ratquad <- abs(theta1-star)/(abs(theta1 - theta0))^2

does diverges. This implies our algorithm is super-linearly convergent.

e.g. the plots show that $\lim_{n \to \infty}\frac{|\theta^{n+1} - \theta^*|}{|\theta^n - \theta^*|} =0$ and $\lim_{n \to \infty}\frac{|\theta^{n+1} - \theta^*|}{|\theta^n - \theta^*|^2} = \infty$

## (d).

```{r}
df.d <- secM2.c(-.2,.2,10,1e-6,1e-9,thetstar)[,3:4]
df.d
plot(1:length(df.d[,1]),df.d[,1],type = 'b')
plot(1:length(df.d[,1]),df.d[,2],type = 'b')

```

When we choose two initial points -0.2 and 0.2 then both of our ratio. 

The plots show that both $\lim_{n \to \infty}\frac{|\theta^{n+1} - \theta^*|}{|\theta^n - \theta^*|} = \infty$

and $\lim_{n \to \infty}\frac{|\theta^{n+1} - \theta^*|}{|\theta^n - \theta^*|^2} = \infty$ implying a divergent secant algorithm. 

